A non-additive generalization of probability theory \\for quantum mechanics and beyond

### 2024-11-07 Articles/2024-NonAdditiveProbability/NonAdditiveProbability.tex
NonAdditiveProbability: reviewed intro

\usepackage{assumptionsofphysics}
\usepackage{tikz}
\usepackage{breakurl}

@@

As mentioned in the introduction, our goal is to construct a probability theory that is guaranteed to work in any physical theory. Therefore our fundamental axioms will not be about probability, but about physical theories. This will help us place probability in a bigger context, and provide a minimal interpretation. As it is common in our larger project, we want axioms to describe necessary requirements of the scientific practice. That is, they will describe features of physical practice, not feature of the systems being described.

We are looking to find a minimal set of axioms required by a physical theory that allows statistical descriptions. Such a theory must, at the very least, provide us with a notion of ensemble, and the space of ensembles that are allowed in the theory. An ensemble can be understood as the infinite collection of the outputs of a particular preparation procedure. The ensemble space of classical mechanics, for example, is the space of continuous probability distributions of phase space. In quantum mechanics, instead, it is the space of density operators. While the ensemble space for each theory will be in principle different, some will have many common features. In all cases we must be able to talk about conservative or dissipative processes, measurable properties of the system, probability distributions for outcomes of said properties and so on. We are looking for necessary minimal basic requirements upon which these common tools can be built.

At least two arguments can be made to show that any physical theory must provide a set of ensembles. First, note that physical laws are never about specific instances but rather regularities. They are statements of the type ``whenever we prepare this, we can measure that.'' This means that, ultimately, they are about ensembles. Second, if a physical theory is to be testable experimentally, and repeatedly so, it must be in terms of ensembles because this is what we prepare in practice and this is what, in the end, we characterized experimentally. It seems, then, that the approach can be given firm conceptual grounds, which we leave as open but solvable philosophical problem.

@@

\subsection{Statistical mixtures}
Another basic requirement is the ability to prepare a \textbf{mixture} of two ensembles. If $\ens[a], \ens[b] \in \Ens$ are two ensembles, and $p \in [0,1]$ a weight, then $\ens = p \ens[a] + (1-p) \ens[b] \in \Ens$ that describes a process that selects the first ensemble over the second $p$ percent of the times. Mathematically, the ensemble space is endowed with a \textbf{convex structure}.

@@

\subsection{Entropy}
The last requirement is that each ensemble must have a well defined entropy. That is, there is a scalar function $S : \Ens \to \mathbb{R}$ that satisfies the following
\begin{enumerate}
	\item strictly convex: $S(p \ens[a] + (1-p) \ens[b]) - (p S(\ens[a]) + (1-p) S(\ens[b]) ) \geq 0$
	\item upper bounded: $S(p \ens[a] + (1-p) \ens[b]) - (p S(\ens[a]) + (1-p) S(\ens[b]) ) \leq - p \log p - (1-p) \log(1-p)$
\end{enumerate}
The first tells us that, during mixing, the entropy cannot decrease, and it stays the same if and only we are mixing an ensemble with itself. The second tells us that the most the entropy can increase is given by the choice between the two ensembles.\footnote{The use of the Shannon entropy for the upper bound can likely be recovered.}

The entropy is ultimately responsible for all geometrical structure in physics. In classical mechanics, the geometry is essentially defined by volumes in phase space. The entropy of uniform distributions is given by the logarithm of the volume, meaning that given the volumes we are able to calculate the entropy and given the entropy we are able to reconstruct the volume. In quantum mechanics, the inner product (i.e. the Born rule) allows us to calculate the entropy of the mixture of two pure states, and given that entropy we can recover the Born rule.

An interesting insight is that the entropy, being strictly concave, has a negative defined Hessian. The negation of the Hessian, then, is a symmetric positive defined function of two infinitesimal variations, and can serve as a metric tensor. In both classical and quantum mechanics this recovers the Fisher-Rao metric.

Another important feature is that if two ensembles saturate the upper bound, then they are orthogonal in both classical and quantum mechanics. Therefore we define two ensembles to be \textbf{orthogonal} if their mixture lead to a maximal entropy increase. If two states are orthogonal, on physical grounds, the must have no common component. However, the reverse is not case: two pure states in quantum mechanics have no components to have in common, but are not necessarily orthogonal. However, classical probability spaces are exactly those space in which no common component implies orthogonality. This is the main source of the ``weirdness'' of quantum mechanics.

@@

Each structure is important, but the balance between them is even more important. The key is to understand, on physical grounds, where each problem lies and therefore use the appropriate tool. For example, one may focus on the convex/linear structure and may be tempted to close the space under infinite convex/linear combinations. Mathematically, it would be very convenient. However, not all infinite mixtures correspond to physically meaningful systems, such as a particle with an infinite average position. It is the connection to experimental verification, the topology, that will tell us which limits are allowed and which are not. This type of conceptual clarity has to come first. The math follows. A full examination of these issues is beyond the scope of this article, and we refer to our larger, and future, work. However, there is one key issue that we need to discuss as, we found, it was the main conceptual obstacle to progress.

On the convex structure we can define \textbf{pure states} as the ones that cannot be decomposed into other ensembles. For example, the pure states for a six-sided die are the ensemble perfectly prepared respectively on the six possible outcomes. All ensembles are convex combinations $\sum_{i+1}^{6} p_i = 1$ of those pure states. Mathematically, the pure states are the extreme points of the space and all other points can be understood as convex combinations over them. This is exactly how classical discrete spaces work. Most physicists, including are older selves, think that that's how everything works. For example, in classical mechanics the pure states are perfect preparations of position and momentum, and ensembles are probability distributions. This does not work.

We cannot consistently prepare classical particles perfectly, therefore the points of phase space are not ensemble. The mathematical facts that delta Dirac are not proper functions, have support with empty interior, have minus infinite entropy, and so on, are simply a consequence of our conceptual problem. A bit better is to think them as the limit of infinitesimal decomposition: classical ensembles can be decomposed into more and more precise ones, and the points of are the limit of this decomposition. Probability distributions are defined on these limits. This works in classical mechanics, but in general.

In quantum mechanics we have pure states, and these are ensembles that cannot be further decomposed into finer ensembles. Therefore an eigenstate of position is not a limit of decomposition, a limit in the convex structure. It is the limit where we are preparing a pure state on a smaller and smaller region of space. It is the limit of a sequence of pure states, a limit in the topology. Physically, more than mathematically, it is a completely different thing, therefore it is not a good route to generalization.\footnote{The fact that pure states can be seen, in some sense, as superposition of eigenstates of position, makes the matter even more confusing.}

The insight is that probability distributions for a quantity are not defined by statistical mixtures of perfect preparations (i.e. a convex combination of extreme points), but by assigning a probability over each experimentally verifiable (i.e. open) interval of the possible values of said quantity (its spectra). The two coincide in classical discrete space, which is the least physically interesting case as it does not even allow for classical particle mechanics.

>>>>

\usepackage{assumptionsofphysics}
\usepackage{tikz}
\usepackage{breakurl}
\usepackage{tcolorbox}

@@

As mentioned in the introduction, our goal is to construct a probability theory of general applicability to physical systems. Our approach is not to define an axiomatic theory of probability and apply to physical systems, but rather characterize basic properties of physical states and from those build a theory of probability that fits those definitions. As it is common in our larger project, we start from axioms that, in our opinion, represent requirements of the scientific practice. The advantage is that the realm of applicability and physical interpretation will be clear from the start.

We start from requiring a physical theory to allow statistical descriptions: at the very least, it will provide us with a notion of ensemble, and the space of ensembles that are allowed by the theory. An ensemble can be understood as the infinite collection of the outputs of a particular preparation procedure. The ensemble space of classical mechanics, for example, is the space of continuous probability distributions of phase space. In quantum mechanics, instead, it is the space of density operators. While the ensemble space for each theory will be in principle different, they will all share common features: they must allow measurable properties of the system, probability distributions for outcomes of said properties, characterize reversible and irreversible processes through an entropy function, and so on. We are looking for necessary minimal basic requirements upon which these common tools can be built.

At least two arguments can be made to show that any physical theory must allow statistical descriptions, provide an ensemble space. First, note that physical laws are never about specific instances but rather regularities. They are statements of the type ``whenever we prepare this, we can measure that.'' This means that, ultimately, they are about ensembles. Second, if a physical theory is to be testable experimentally, and repeatedly so, it must be in terms of ensembles because this is what we prepare in practice and this is what, in the end, we characterize experimentally. The approach, then, can most likely be given firm conceptual grounds, which we leave as open but solvable philosophical problem.

@@

In the case of ensembles, typical verifiable statements will be ``the average energy of the particle is between 3 and 4 $eV$,'' ``the probability of getting heads is between 49 and 51 percent,'' ``the probability distribution for the position is a Gaussian of average 0 $m$ and 1 $m$ standard deviation with 1\%.'' Note that these are not results of single shot measurements, and therefore verifiable statements are distinctly different from quantum measurements. In the same manner, these are not measurements that extract one bit of information, as there is no requirement of termination in the negative case.

\begin{tcolorbox}[colback=white, colframe=black]
	Experimental verifiability $\Rightarrow$ An ensemble space $\Ens$ is a $T_0$ second-countable topological space.
\end{tcolorbox}

\subsection{Statistical mixtures}
Another basic requirement is the ability to prepare a \textbf{mixture} of two ensembles. If $\ens[a], \ens[b] \in \Ens$ are two ensembles, and $p \in [0,1]$ a weight, then $\ens = p \ens[a] + (1-p) \ens[b] \in \Ens$ is the ensemble that describes a process that selects the first ensemble over the second $p$ percent of the times. Mathematically, the ensemble space is endowed with a \textbf{convex structure}. Moreover, since the mixing operation must be consistent with experimental verifiability, it will be topologically continuous an the ensemble space will be a topological convex space.

@@

\begin{tcolorbox}[colback=white, colframe=black]
	Statistical mixture of ensemble $\Rightarrow$ An ensemble space $\Ens$ is a topological convex space.
\end{tcolorbox}

\subsection{Entropy}
The last requirement is that each ensemble must have a well defined \textbf{entropy}. That is, there is a scalar function $S : \Ens \to \mathbb{R}$ that satisfies the following
\begin{enumerate}
	\item strictly convex: $S(p \ens[a] + (1-p) \ens[b]) - (p S(\ens[a]) + (1-p) S(\ens[b]) ) \geq 0$
	\item upper bounded: $S(p \ens[a] + (1-p) \ens[b]) - (p S(\ens[a]) + (1-p) S(\ens[b]) ) \leq - p \log p - (1-p) \log(1-p)$
\end{enumerate}
The first tells us that, during mixing, the entropy cannot decrease, and it stays the same if and only we are mixing an ensemble with itself. The second tells us that the most the entropy can increase is given by the choice between the two ensembles.\footnote{It can be shown that if an upper bound that only depends on the mixing coefficients exists, then it must be the Shannon entropy.}

The entropy is ultimately responsible for all geometrical structure in physics. In classical mechanics, the geometry is essentially defined by phase space volumes and areas in each DOF (degree of freedom). The entropy of uniform distributions is given by the logarithm of the volume, meaning that given the volumes we are able to calculate the entropy and given the entropy we are able to reconstruct the volume. Similarly, uniform marginals will recover DOF areas. In quantum mechanics, the inner product (i.e. the Born rule) allows us to calculate the entropy of the mixture of two pure states, and given the entropy we can recover the Born rule.

An interesting insight is that the entropy, being strictly concave, has a negative defined Hessian. The negation of the Hessian, then, is a symmetric positive defined function of two infinitesimal variations, and can serve as a metric tensor on the affine structure given by the mixing coefficients. In both classical and quantum mechanics this recovers the Fisher-Rao metric.

Another important feature is that if two ensembles saturate the upper bound, then they are orthogonal in both classical and quantum mechanics. Therefore we define two ensembles to be \textbf{orthogonal} if their mixture lead to a maximal entropy increase. An additional axiom is valid for orthogonal ensembles. We say an ensemble is a component of another ensemble if the second can be expressed as a mixture of the first with some other ensemble. On physical grounds, two orthogonal ensembles $\ens_1, \ens_2 \in \Ens$ are perfectly distinguishable and must have no common component. That is, there is no $\ens_3 \in \Ens$ that is a component of both $\ens_1$ and $\ens_2$. Note that the converse is not generally true: two pure states in quantum mechanics have no components to have in common, but are not necessarily orthogonal. We find that an ensemble space where no common component implies orthogonality can be described by a space of classical probability measures. Non-classicality, then, is exactly the failure of this property.

\begin{tcolorbox}[colback=white, colframe=black]
	Existence of entropy $\Rightarrow$ Strictly concave function whose upper bound defines orthogonality.
\end{tcolorbox}

@@

Each structure is important, but the balance between them is even more important. The key is to understand, on physical grounds, where each problem lies and therefore use the appropriate tool. For example, one may focus on the convex/linear structure and may be tempted to close the space under infinite convex/linear combinations. Mathematically, it would be very convenient. However, some infinite mixtures lead to physically ill-defined systems, such as a particle with an infinite average position. It is the connection to experimental verification, the topology, that will tell us which limits are allowed and which are not. Conceptual clarity on these details is essential to find the correct generalization. A full examination of these issues is beyond the scope of this article, and we refer to our larger, and future, work. However, there is one key issue that we need to discuss as, we found, it was the main conceptual obstacle to progress.

On the convex structure we can define \textbf{pure states} as the ones that cannot be decomposed into other ensembles. For example, the pure states for a six-sided die are the ensemble perfectly prepared respectively on the six possible outcomes. All ensembles are convex combinations $\sum_{i=1}^{6} p_i \ens_i = 1$ of those pure states $\ens_i$. Mathematically, the pure states are the extreme points of the space and all other points can be understood as convex combinations over them. This is exactly how classical discrete spaces work and there is a temptation to think this is how everything works. For example, in classical mechanics the pure states would be perfect preparations of position and momentum, and ensembles are probability distributions. Not only this is problematic mathematically, it is conceptually not sustainable.

On physical grounds, we cannot consistently prepare classical particles perfectly, therefore there is no ensemble corresponding to a perfect preparation of a point of phase space. The mathematical facts that delta Dirac are not proper functions, their support has an empty interior, have minus infinite entropy, and so on, are simply a consequence of our conceptual problem. One may concede that the ensembles to do not exist physically, but may still want them to consider them as the limit of infinitesimal decomposition: classical ensembles can be decomposed into more and more precise ones, and the points of are the limit of this decomposition. Probability distributions are defined on these limits. This kind of works in classical mechanics, but in general.

In quantum mechanics we do have pure states, and these are ensembles that cannot be further decomposed into finer ensembles. Therefore an eigenstate of position is not a limit of decomposition, a limit in the convex structure. It is the limit where we are preparing a pure state on a smaller and smaller region of space. It is the limit of a sequence of pure states, a limit in the topology. Physically, more than mathematically, it is a completely different thing. It is only when we realized this difference that a clean route to generalization presented itself.\footnote{The fact that pure states can be seen, in some sense, as superposition of eigenstates of position, makes the matter even more confusing.}

The insight is that probability distributions for a quantity are not defined by statistical mixtures of perfect preparations (i.e. a convex combination of extreme points), but by assigning a probability over each experimentally verifiable (i.e. open) interval of the possible values of said quantity (its spectra). The two coincide in classical discrete spaces, which are the least physically interesting case as they do not even allow classical particle mechanics. Equivalently, \textbf{mixtures are about preparations, not measurements, and probability distributions are about measurements, not preparations}. Since they can only coincide in classical spaces, a generalized notion of probability for physical theories must resolve and characterize the distinction.
